---
title: 'Variable selection in topic models: Alex Bird Data'
author: "Kushal K Dey"
date: "December 27, 2015"
output: 
  html_document:
    toc: true
---

## Background

We apply the variable selection method on real data. We first focus on the bird abundance data in Himalayas, which we are collaborating on with Alex White and Trevor Price. 

## Data Pre-processing

```{r echo=TRUE, eval=TRUE}

data=read.csv("../external_data/Himalayan_grid_matrix.csv",header=TRUE);

counts=as.matrix(data[,-1]);

rownames(counts)=data[,1];

new_data1 <- data.frame(read.csv('../external_data/MohanGrids.csv'));
new_data2 <- data.frame(read.csv('../external_data/MohanGrids2.csv'));

bird_species <- union(as.vector(colnames(counts)), union(as.vector(new_data1[,1]), as.vector(new_data2[,1])));

new_data <- matrix(0,dim(counts)[1]+3,length(bird_species));
new_data[1,match(new_data1[,1],bird_species)]=new_data1[,2];
new_data[2,match(new_data1[,1],bird_species)]=new_data1[,3];
new_data[3,match(new_data2[,1],bird_species)]=new_data2[,2];
new_data[4:(dim(counts)[1]+3),match(colnames(counts),bird_species)]=counts;

new_counts <- as.matrix(new_data);

rownames(new_counts) <- c(c("U1","U2","MA1"),rownames(counts));
colnames(new_counts) <- bird_species;
new_counts <- new_counts[-(1:3),];

```

## Topic model without variable selection

We first apply the topic model without variable selection pre-processing.

```{r echo=TRUE, eval=TRUE}
metadata=read.csv("../external_data/Himalayan_grid_metadata.csv",header=TRUE);
elevation_metadata=metadata$Elevation[match(rownames(new_counts),metadata[,1])];

east_west_dir = metadata$WorE[match(rownames(new_counts),metadata[,1])];

library(maptpx)
system.time(Topic_clus <- topics(new_counts, K=2,tol=0.01));
K=2
barplot(t(Topic_clus$omega[order(elevation_metadata),]),col=2:(K+1),axisnames=F,space=0,border=NA,main="",las=1,ylim=c(0,1),cex.axis=1.5,cex.main=1.4)
title(main=paste("Structure Plot topic proportions,k=",K))
combo_patch_dir = paste0(elevation_metadata,"_",east_west_dir);
combo_patch_dir_ordered = combo_patch_dir[order(elevation_metadata)];

match_labs=match(unique(combo_patch_dir_ordered),combo_patch_dir_ordered);
match_labs_suffix=c(match_labs[2:length(unique(combo_patch_dir_ordered))],35);
match_labs_prefix=match_labs[1:(length(unique(combo_patch_dir_ordered)))];
labs=match_labs_prefix + 0.5*(match_labs_suffix - match_labs_prefix);
  
axis(1,at=labs,unique(combo_patch_dir_ordered),las=2);
abline(v=match_labs[2:length(match_labs)]);

```

## Topic model with variable selection

### Mean centering

We mean center the rows and the columns.

```{r echo=TRUE, eval=TRUE}
data_norm1 <- apply(new_counts,2,function(x) return(x-mean(x)));
data_norm2 <- apply(data_norm1, 1, function(x) return(x-mean(x)));
```

### Penalized matrix decomposition

Then we perform factor analysis on the genes.

```{r echo=TRUE, eval=TRUE}
library(PMA)
pmd1 <- PMD(data_norm2, type="standard", K=2, niter=50);
```

### Factor loadings explore

We plot the barplot of the factor loadings, colors representing the two factors.

```{r echo=TRUE, eval=TRUE}
pos_u <- apply(pmd1$u, c(1,2), function(x) return (max(x,0)))
neg_u <- apply(pmd1$u, c(1,2), function(x) return (min(x,0)))

par(mar=c(12,2,2,1))
barplot(t(pos_u), col=2:(K+1))
barplot(t(neg_u), col=2:(K+1), add=TRUE)

```

### Shrinking factor loadings

The sum of squared values of the columns of $u$ matrix from *PMD()* fit is 1 by definition. Check here 

```{r echo=TRUE, eval=TRUE}
apply(pmd1$u,2,function(x) return(sqrt(sum(x^2))))
```

We assume that all the genes contribute equally to this variance or standard deviation of $1$ in the $u$ matrix and we assume the variance to be $1/G$, whhere $G$ is the total number of genes.

```{r echo=TRUE, eval=TRUE}
library(ashr)
shrunk_u <- suppressWarnings(apply(pmd1$u, 2, function(x) return(ash(x,sqrt(1/length(x)))$PosteriorMean)));
pos_shrunk_u <- apply(shrunk_u, c(1,2), function(x) return (max(x,0)))
neg_shrunk_u <- apply(shrunk_u, c(1,2), function(x) return (min(x,0)))

par(mar=c(12,2,2,1))
barplot(t(pos_shrunk_u), col=2:3)
barplot(t(neg_shrunk_u), col=2:3, add=TRUE)

```

We compute the sum of squares of the loadings for the shrunken $u$ matrix. We choose the threshold to be $0.001$ for the loadings.

### Topic model fit on selected species (threshold 0.0001)

```{r echo=TRUE, eval=TRUE}
ss_loadings <- apply(shrunk_u,1,function(x) return(sqrt(sum(x^2))))
which(ss_loadings > 1e-04)
```

We now perform topic model on the selected genes.

```{r echo=TRUE, eval=TRUE}
vs_counts <- new_counts[,which(ss_loadings > 1e-04)];
system.time(Topic_clus <- topics(vs_counts, K=2,tol=0.01));
K <- 2;
barplot(t(Topic_clus$omega[order(elevation_metadata),]),col=2:(K+1),axisnames=F,space=0,border=NA,main="",las=1,ylim=c(0,1),cex.axis=1.5,cex.main=1.4)
title(main=paste("Structure Plot topic proportions,k=",K))
combo_patch_dir = paste0(elevation_metadata,"_",east_west_dir);
combo_patch_dir_ordered = combo_patch_dir[order(elevation_metadata)];

match_labs=match(unique(combo_patch_dir_ordered),combo_patch_dir_ordered);
match_labs_suffix=c(match_labs[2:length(unique(combo_patch_dir_ordered))],35);
match_labs_prefix=match_labs[1:(length(unique(combo_patch_dir_ordered)))];
labs=match_labs_prefix + 0.5*(match_labs_suffix - match_labs_prefix);
  
axis(1,at=labs,unique(combo_patch_dir_ordered),las=2);
abline(v=match_labs[2:length(match_labs)]);

```


### Topic model fit on selected species (threshold 0.00001)


```{r echo=TRUE, eval=TRUE}
ss_loadings <- apply(shrunk_u,1,function(x) return(sqrt(sum(x^2))))
which(ss_loadings > 1e-05)
```

We now perform topic model on the selected genes.

```{r echo=TRUE, eval=TRUE}
vs_counts <- new_counts[,which(ss_loadings > 1e-05)];
system.time(Topic_clus <- topics(vs_counts, K=2,tol=0.01));
K <- 2;
barplot(t(Topic_clus$omega[order(elevation_metadata),]),col=2:(K+1),axisnames=F,space=0,border=NA,main="",las=1,ylim=c(0,1),cex.axis=1.5,cex.main=1.4)
title(main=paste("Structure Plot topic proportions,k=",K))
combo_patch_dir = paste0(elevation_metadata,"_",east_west_dir);
combo_patch_dir_ordered = combo_patch_dir[order(elevation_metadata)];

match_labs=match(unique(combo_patch_dir_ordered),combo_patch_dir_ordered);
match_labs_suffix=c(match_labs[2:length(unique(combo_patch_dir_ordered))],35);
match_labs_prefix=match_labs[1:(length(unique(combo_patch_dir_ordered)))];
labs=match_labs_prefix + 0.5*(match_labs_suffix - match_labs_prefix);
  
axis(1,at=labs,unique(combo_patch_dir_ordered),las=2);
abline(v=match_labs[2:length(match_labs)]);
```

## Conclusion

The Structure plots after variable selection are slightly different from that without variable selection. Note that for threshold $0.0001$, we get a smooth continuous pattern in cluster memberships in the Structure plot. But that pattern is not seen for threshold $0.00001$, probably due to some false species been taken into account. We need to fix the threshold selection issue first. On a broad scale, the main patterns remain preserved from the Structure plot without variable selection to the one with variable selection.

