---
title: 'Wavelets: applying normal topic model on sinusoidal patterns'
author: "Kushal K Dey"
date: "December 13, 2015"
output: 
  html_document:
    toc: true
---

## Background

In this script, we apply the Normal Topic Model (NTM) on sinusoidally varying gene expression data matrix with and without wavelet transformation. The idea is to replicate patterns similar to sparse factor analysis, but to focus more on clustering the samples based on the expression patterns.

```{r echo=TRUE, eval=TRUE}
source('../R/ntmfit.R')
```

## Toy example (sinusoidal data matrix)

We simulate the sinusoidal expression for $100$ genes over $256$ time points in order, where the amplitudes for all the genes are $10$, the phases for the first $50$ genes are $2$ and that of the last $50$ genes being $5$. The error variation besides the signal for each genes is
taken to be generated from $\chi^2_{4}$ distribution. So, there are two groups of genes depending on gene phases (the first $50$ coming from one phase and the last $50$ from another).

```{r echo=TRUE, eval=TRUE}
library(PMA)
library(cellcycleR)
library(wavethresh)
G <- 100;
num_cells <- 256;
amp_genes1 <- rep(10, G);
phi_genes1 <- rep(c(2,5), each=G/2);
sigma_genes1 <- rchisq(G, 4);
cell_times_sim <- sort(sample(seq(0,2*pi, 2*pi/(num_cells-1)), num_cells, replace=FALSE));
cycle_data <- t(sim_sinusoidal_cycle(G, amp_genes1, phi_genes1, sigma_genes1, cell_times_sim));

```

The example plots from above example

```{r echo=TRUE, eval=TRUE}
plot(cycle_data[2,], type="l");
plot(cycle_data[(2+G/2), ], type="l")
```

## Applying Normal Topic Model 

Next, we apply the *NTMfit()* function on the this data.

```{r echo=TRUE, eval=TRUE}
system.time(out1 <- NTMfit(cycle_data, K=2));
```

We plot the Structure plot for the topic proportions obtained from the model fit.

```{r echo=TRUE, eval=TRUE}
K <- 2;
barplot(t(out1$omega_out),col=2:(K+1),axisnames=F,space=0,border=NA,main="",las=1,ylim=c(0,1),cex.axis=1.5,cex.main=1.4)
title(main=paste("Structure Plot topic proportions,k=",K))

plot(out1$alpha_out[1,], type="l")
plot(out1$alpha_out[2,], type="l")

```

## Wavelet Transformation of data matrix

```{r echo=TRUE, eval=TRUE}

wave_data <- t(apply(cycle_data, 1, function(x)
                                  {
                                      out <- wd(x, filter.number = 3, family = "DaubExPhase")
                                      return(out$D)
                                  }));
```

## Applying Normal Topic Model on wavelet transformed data

Now apply *NTMfit()* function on the this wavelet-transformed data.

```{r echo=TRUE, eval=TRUE}
system.time(out2 <- NTMfit(wave_data, K=2));
```


We plot the Structure plot for the topic proportions obtained from the above model fit.

```{r echo=TRUE, eval=TRUE}
K <- 2;
barplot(t(out2$omega_out),col=2:(K+1),axisnames=F,space=0,border=NA,main="",las=1,ylim=c(0,1),cex.axis=1.5,cex.main=1.4)
title(main=paste("Structure Plot topic proportions,k=",K))

```

## Recreating back factors from wavelet transform

Reforming the cluster patterns by reverse transforming the *alpha* output from the *NTMfit()* model fit above.

```{r echo=TRUE, eval=TRUE}
W <- GenW(n=num_cells, filter.number=3, family="DaubExPhase")
idwt <- W[,-1] %*% t(out2$alpha_out);
plot(idwt[,1], type="l")
plot(idwt[,2], type="l")

```

## Conclusions

The NTM (Normal Topic model) does a pretty decent job at clustering the two groups of sinusoidal wave patterns both without and with wavelet transforms.

