---
title: 'Voom topic models: PMD application 2'
author: "Kushal K Dey"
date: "February 12, 2016"
output: html_document
---

## Overview

We perform a penalized matrix decomposition on transformations of the counts data. We use three types of transformations.

- Square root transform
- Log cpm transform

## Simulated design

We first cosider a simulated design model.

```{r echo=TRUE, eval=TRUE}
n.out <- 200
omega_sim <- cbind(seq(0.6,0.4,length.out=n.out), 1- seq(0.6,0.4,length.out=n.out));
K <- dim(omega_sim)[2];
barplot(t(omega_sim),col=2:(K+1),axisnames=F,space=0,border=NA,main=paste("No. of clusters=",K),las=1,ylim=c(0,1),cex.axis=1.5,cex.main=1.4)
freq <- rbind(c(0.1,0.2,rep(0.70/98,98)),c(rep(0.70/98,98), 0.1,0.2));
counts <- t(do.call(cbind,lapply(1:dim(omega_sim)[1], function(x) rmultinom(1,1000,prob=omega_sim[x,]%*%freq))));
```

We transform the counts data 

### voom transform

```{r echo=TRUE, eval=TRUE}
voom_data <- limma::voom(counts)$E;
```

### PMD transform 

```{r echo=TRUE, eval=TRUE}
library(PMA)
voom_data_norm <- apply(voom_data, 2, function(x) (x - mean(x)))
voom_data_norm_norm <- apply(voom_data_norm, 1, function(x) (x - mean(x)))
```

```{r echo=TRUE, eval=TRUE}
pmd_out <-PMD(voom_data_norm_norm, K=2, vpos=TRUE, upos=TRUE, niter=500, sumabsv=2);
```

```{r echo=TRUE, eval=FALSE}
pmd_out <-PMD(voom_data, K=2, niter=500, sumabsv=2);
```

### Plotting the u from pmd

```{r echo=TRUE, eval=TRUE}
loadings <- pmd_out$u
```

```{r echo=TRUE, eval=TRUE}
barplot(t(loadings),col=2:(K+1),axisnames=F,space=0,border=NA,main=paste("Structure Plot, K=",K),las=1,cex.axis=1.5,cex.main=1.4)
```

### Plotting the v from pmd

```{r echo=TRUE, eval=TRUE}
factor_dist <- maptpx::normalize(pmd_out$v, byrow=FALSE);
barplot(t(factor_dist),col=2:(K+1),axisnames=F,space=0,border=NA,main=paste("Factor distribution, K=",K),las=1,cex.axis=1.5,cex.main=1.4)
```

Now we use the square root transform.

```{r echo=TRUE, eval=TRUE}
sqrt_data <- sqrt(counts);
sqrt_data_norm <- apply(sqrt_data, 2, function(x) (x - mean(x)))
sqrt_data_norm_norm <- apply(sqrt_data_norm, 1, function(x) (x - mean(x)))
```

```{r echo=TRUE, eval=TRUE}
pmd_out <-PMD(sqrt_data_norm_norm, K=2, sumabsv=2);
```

### Plotting the u from pmd

```{r echo=TRUE, eval=TRUE}
loadings <- pmd_out$u
```

```{r echo=TRUE, eval=TRUE}
barplot(t(loadings),col=2:(K+1),axisnames=F,space=0,border=NA,main=paste("Structure Plot, K=",K),las=1,cex.axis=1.5,cex.main=1.4)
```

### Plotting the v from pmd

```{r echo=TRUE, eval=TRUE}
factor_dist <- maptpx::normalize(pmd_out$v, byrow=FALSE);
barplot(t(factor_dist),col=2:(K+1),axisnames=F,space=0,border=NA,main=paste("Factor distribution, K=",K),las=1,cex.axis=1.5,cex.main=1.4)
```


## Transposing the data and re-doing PMD (voom transform)

Next we performed PMD() on the transposed dat. In this case, the loadings will be on the genes and the factors would represent weights on the different samples.

```{r echo=TRUE, eval=TRUE}
pmd_out <-PMD(t(voom_data_norm_norm), K=2, sumabsu=0.3);
```

### Plotting the u from pmd

```{r echo=TRUE, eval=TRUE}
loadings <- pmd_out$u
```

```{r echo=TRUE, eval=TRUE}
barplot(t(loadings),col=2:(K+1),axisnames=F,space=0,border=NA,main=paste("Facror distribution, K=",K),las=1,cex.axis=1.5,cex.main=1.4)
```

### Plotting the v from pmd

```{r echo=TRUE, eval=TRUE}
factor_dist <- maptpx::normalize(pmd_out$v);
barplot(t(factor_dist),col=2:(K+1),axisnames=F,space=0,border=NA,main=paste("Structure Plot, K=",K),las=1,ylim=c(0,1),cex.axis=1.5,cex.main=1.4)
```

## Applying without transformation

```{r echo=TRUE, eval=TRUE}
counts_norm <- apply(counts, 2, function(x) (x - mean(x)));
counts_norm_norm <- apply(counts_norm, 1, function(x) (x - mean(x)));
pmd_out <-PMD(counts_norm_norm, K=3, vpos=TRUE, upos=TRUE, niter=500);
```

### Plotting the u from pmd

```{r echo=TRUE, eval=TRUE}
loadings <- pmd_out$u
```

```{r echo=TRUE, eval=TRUE}
K <- 2
barplot(t(loadings),col=2:(K+1),axisnames=F,space=0,border=NA,main=paste("Structure Plot, K=",K),las=1,ylim=c(0,1),cex.axis=1.5,cex.main=1.4)
```

### Plotting the v from pmd

```{r echo=TRUE, eval=TRUE}
factor_dist <- maptpx::normalize(pmd_out$v, byrow=FALSE);
barplot(t(factor_dist),col=2:(K+1),axisnames=F,space=0,border=NA,main=paste("Factor distribution, K=",K),las=1,cex.axis=1.5,cex.main=1.4)
```

