---
title: 'Variable selection in topic models: (K=3)'
author: "Kushal K Dey"
date: "December 26, 2015"
output: 
  html_document:
    toc: true
---

## Overview

In this script, we extend the variable selection mechanism from $K=2$ to $K=3$.  We use the *PMD()* function of the R package **PMA**  for factor analysis and then use the factor loadings matrix to do the variable selection.

## Simulation design

```{r echo=TRUE, eval=TRUE}
library(maptpx)
n.out <- 200
omega_sim <- cbind(seq(0.4,0.2,length.out=n.out), seq(0.2,0.4,length.out=n.out), 1- seq(0.4,0.2,length.out=n.out)-seq(0.2,0.4,length.out=n.out));
K <- dim(omega_sim)[2];
barplot(t(omega_sim),col=2:(K+1),axisnames=F,space=0,border=NA,main=paste("No. of clusters=",K),las=1,ylim=c(0,1),cex.axis=1.5,cex.main=1.4)
freq <- rbind(c(0.1,0.2,rep(0.70/98,98)),c(rep(0.70/98,98), 0.1,0.2), c(0.1,0.1,rep(0.4/94,47),0.1,0.1, rep(0.4/94,47),0.1,0.1));
counts <- t(do.call(cbind,lapply(1:dim(omega_sim)[1], function(x) rmultinom(1,1000,prob=omega_sim[x,]%*%freq))));

```

By the way we have defined the set up, besides the first two, $50$, $51$ and the last two genes, all the other genes are unimportant, or in other words, all the essential information required for the clustering is contained in the above genes. Ideally, we would like to eliminate the non-informative genes before applying the topic model for more accurate representation, removing unwanted noise and also faster implementation.

## Topic model without variable selection

We first apply the topic model without variable selection pre-processing.

```{r echo=TRUE, eval=TRUE}
system.time(Topic_clus <- topics(counts, K=3,tol=0.001));
K=3
docweights <- Topic_clus$omega;
library(permute);
library("BioPhysConnectoR");
perm_set=rbind(1:K,allPerms(1:K));
diff=array(0,dim(perm_set)[1]);
for (p in 1:dim(perm_set)[1])
{
	temp=docweights[,perm_set[p,]];
	diff[p]=fnorm(temp,omega_sim);
}

p_star=which(diff==min(diff));
docweights=docweights[,perm_set[p_star,]];

barplot(t(docweights),col=2:(K+1),axisnames=F,space=0,border=NA,main=paste("No. of clusters=",K),las=1,ylim=c(0,1),cex.axis=1.5,cex.main=1.4)

```

## Factor analysis on data

Now we perform the variable selection. Before applying the factor analysis model, we first mean correct the columns and rows. 

```{r echo=TRUE, eval=TRUE}
data_norm1 <- apply(counts,2,function(x) return(x-mean(x)));
data_norm2 <- apply(data_norm1, 1, function(x) return(x-mean(x)));
```

Then we perform factor analysis on the genes.

```{r echo=TRUE, eval=TRUE}
library(PMA)
pmd1 <- PMD(data_norm2, type="standard", K=3, niter=50);
```

We plot the barplot of the factor loadings, colors representing the two factors.

```{r echo=TRUE, eval=TRUE}
pos_u <- apply(pmd1$u, c(1,2), function(x) return (max(x,0)))
neg_u <- apply(pmd1$u, c(1,2), function(x) return (min(x,0)))

par(mar=c(12,2,2,1))
barplot(t(pos_u), col=2:(K+1))
barplot(t(neg_u), col=2:(K+1), add=TRUE)

```

## Shrinking factor loadings

From the barplot, it is clear that mainly $6$ genes have high loadings and are informative about clustering. So, these genes need to be kept in the clustering. However, the other genes which are non-informative still have small loadings. We use shrinkage to shrink these small values to 0.

The sum of squared values of the columns of $u$ matrix from *PMD()* fit is 1 by definition. Check here 

```{r echo=TRUE, eval=TRUE}
apply(pmd1$u,2,function(x) return(sqrt(sum(x^2))))
```

We assume that all the genes contribute equally to this variance or standard deviation of $1$ in the $u$ matrix and we assume the variance to be $1/G$, whhere $G$ is the total number of genes.

```{r echo=TRUE, eval=TRUE}
library(ashr)
shrunk_u <- suppressWarnings(apply(pmd1$u, 2, function(x) return(ash(x,sqrt(1/length(x)))$PosteriorMean)));
pos_shrunk_u <- apply(shrunk_u, c(1,2), function(x) return (max(x,0)))
neg_shrunk_u <- apply(shrunk_u, c(1,2), function(x) return (min(x,0)))

par(mar=c(12,2,2,1))
barplot(t(pos_shrunk_u), col=2:(K+1))
barplot(t(neg_shrunk_u), col=2:(K+1), add=TRUE)

```

We compute the sum of squares of the loadings for the shrunken $u$ matrix. We choose the threshold to be $0.001$ for the loadings for choosing the genes. 

## Topic model after variable selection (thresh=0.001)

```{r echo=TRUE, eval=TRUE}
ss_loadings <- apply(shrunk_u,1,function(x) return(sqrt(sum(x^2))))
which(ss_loadings > 1e-03)
```

We now perform topic model on the selected genes.

```{r echo=TRUE, eval=TRUE}
vs_counts <- counts[,which(ss_loadings > 1e-03)];
system.time(Topic_clus <- topics(vs_counts, K=3,tol=0.001));
K=3
docweights <- Topic_clus$omega;
library(permute);
library("BioPhysConnectoR");
perm_set=rbind(1:K,allPerms(1:K));
diff=array(0,dim(perm_set)[1]);
for (p in 1:dim(perm_set)[1])
{
	temp=docweights[,perm_set[p,]];
	diff[p]=fnorm(temp,omega_sim);
}

p_star=which(diff==min(diff));
docweights=docweights[,perm_set[p_star,]];

barplot(t(docweights),col=2:(K+1),axisnames=F,space=0,border=NA,main=paste("No. of clusters=",K),las=1,ylim=c(0,1),cex.axis=1.5,cex.main=1.4)

```

## Topic model after variable selection (thresh=0.0001)

```{r echo=TRUE, eval=TRUE}
ss_loadings <- apply(shrunk_u,1,function(x) return(sqrt(sum(x^2))))
which(ss_loadings > 1e-04)
```

We now perform topic model on the selected genes.

```{r echo=TRUE, eval=TRUE}
vs_counts <- counts[,which(ss_loadings > 1e-04)];
system.time(Topic_clus <- topics(vs_counts, K=3,tol=0.001));
K=3
docweights <- Topic_clus$omega;
library(permute);
library("BioPhysConnectoR");
perm_set=rbind(1:K,allPerms(1:K));
diff=array(0,dim(perm_set)[1]);
for (p in 1:dim(perm_set)[1])
{
	temp=docweights[,perm_set[p,]];
	diff[p]=fnorm(temp,omega_sim);
}

p_star=which(diff==min(diff));
docweights=docweights[,perm_set[p_star,]];

barplot(t(docweights),col=2:(K+1),axisnames=F,space=0,border=NA,main=paste("No. of clusters=",K),las=1,ylim=c(0,1),cex.axis=1.5,cex.main=1.4)

```

## Conclusion

Note that the topic model after variable selection gives pretty much similar structure plot representation and also at a much quicker time.
However choosing the right threshold is very important. For threshold $0.001$, we miss out on the genes $50$ and $51$, which are retained along with a false gene $97$ for threshold $0.0001$. As a result of this, the Structure plot for the threshold $0.0001$ is closer to the true Structure plot compared to the one for threshold $0.001$. We would want to choose our threshold appropriately so that we do not lose important genes, at the same time we do not pull in too many false genes.


