---
title: 'Wavelets: applying SFA and NTM on yeast (elu) data'
author: "Kushal K Dey"
date: "December 13, 2015"
output: 
  html_document:
    toc: true
---

## Background

We consider the synchronized yeast cell expression across time in Botstein [paper](http://www.ncbi.nlm.nih.gov/pmc/articles/PMC25624/). 
The data was collected from [Yeast Protein Database](http://quest7.proteome.com/YPDhome.html). We focus here only on the **elu** data, which showed bumpy patterns in **cellcycle** applications. See [here](http://jhsiao999.github.io/singleCell-method/project/analysis/yeast_cellcycleR.html). We apply the penalized matrix decomposition (PMD) and the Normal Topic Model (NTM) with and without wavelet transformation on the elu data.


## Data pre-processing

```{r echo=TRUE, eval=TRUE}
data <- read.table("../external_data/Botstein_data/botstein_cdc.txt", sep="\t", header=TRUE, fill=TRUE);
cdc_data <- data[,grep("elu",colnames(data))];

fillNAgaps <- function(x, firstBack=FALSE) {
    ## NA's in a vector or factor are replaced with last non-NA values
    ## If firstBack is TRUE, it will fill in leading NA's with the first
    ## non-NA value. If FALSE, it will not change leading NA's.
    
    # If it's a factor, store the level labels and convert to integer
    lvls <- NULL
    if (is.factor(x)) {
        lvls <- levels(x)
        x    <- as.integer(x)
    }
 
    goodIdx <- !is.na(x)
 
    # These are the non-NA values from x only
    # Add a leading NA or take the first good value, depending on firstBack   
    if (firstBack)   goodVals <- c(x[goodIdx][1], x[goodIdx])
    else             goodVals <- c(NA,            x[goodIdx])

    # Fill the indices of the output vector with the indices pulled from
    # these offsets of goodVals. Add 1 to avoid indexing to zero.
    fillIdx <- cumsum(goodIdx)+1
    
    x <- goodVals[fillIdx]

    # If it was originally a factor, convert it back
    if (!is.null(lvls)) {
        x <- factor(x, levels=seq_along(lvls), labels=lvls)
    }

    x
}

cdc_data <- cdc_data[rowSums(is.na(cdc_data)) < (dim(cdc_data)[2] - 4),];
cdc_data_mod <- t(apply(cdc_data, 1, function(x) fillNAgaps(x, firstBack = TRUE)));

cycle_data <- cdc_data_mod;
dim(cycle_data)
```

Center and scale the data.

```{r echo=TRUE, eval=TRUE}
cycle_data_norm <- apply(cycle_data,1,function(x)  return (x-mean(x))/sd(x))
cycle_data_norm <- cycle_data_norm[, -which(colSums(cycle_data_norm)==0)]

dim(cycle_data_norm)

```

Padding with $0$ values.

```{r echo=TRUE, eval=TRUE}
padded_data <- t(apply(cycle_data_norm, 2, function(x)
                                             {
                                                  y=x;
                                                  length(y)<-32;
                                                  y[is.na(y)]<-0
                                                  return(y)
                                             }))

```

## Applying PMD on the elu data

Now we first apply the *PMD()* function of the **PMA** package.

```{r echo=TRUE, eval=TRUE}
library(PMA)
pmd1 <- PMD(padded_data, type="standard", K=2, niter=50);


pos1<- apply(pmd1$v, c(1,2), function(x) return (max(x,0)))
neg1 <- apply(pmd1$v, c(1,2), function(x) return (min(x,0)))

par(mar=c(12,2,2,1))
barplot(t(pos1), col=2:3)
barplot(t(neg1), col=2:3, add=TRUE)

plot(pmd1$v[,1], type="l")
plot(pmd1$v[,2], type="l")

pos2 <- apply(pmd1$u, c(1,2), function(x) return (max(x,0)))
neg2 <- apply(pmd1$u, c(1,2), function(x) return (min(x,0)))

par(mar=c(12,2,2,1))
barplot(t(pos2), col=2:3)
barplot(t(neg2), col=2:3, add=TRUE)

```

## Apply Normal Topic model on elu data

Now we apply the NTM model

```{r echo=TRUE, eval=TRUE}
source('../R/ntmfit.R')
system.time(out1 <- NTMfit(padded_data, K=2));
K <- 2;
barplot(t(out1$omega_out),col=2:(K+1),axisnames=F,space=0,border=NA,main="",las=1,ylim=c(0,1),cex.axis=1.5,cex.main=1.4)

plot(out1$alpha_out[1,], type="l")
plot(out1$alpha_out[2,], type="l")
```

## Applying wavelet transformation

We now compute the wavelet transformation of the data matrix.

```{r echo=TRUE, eval=TRUE}
library(wavethresh)
wave_data <- t(apply(padded_data, 1, function(x)
                                  {
                                      out <- wd(x, filter.number = 3, family = "DaubExPhase")
                                      return(out$D)
                                  }));


```

## Applying PMD on wavelet transformed data

We apply factor analysis on the wave-transformed data.

```{r echo=TRUE, eval=TRUE}
library(PMA)
pmd1 <- PMD(wave_data, type="standard", K=2, niter=50);


pos1<- apply(pmd1$v, c(1,2), function(x) return (max(x,0)))
neg1 <- apply(pmd1$v, c(1,2), function(x) return (min(x,0)))

par(mar=c(12,2,2,1))
barplot(t(pos1), col=2:3)
barplot(t(neg1), col=2:3, add=TRUE)

plot(pmd1$v[,1], type="l")
plot(pmd1$v[,2], type="l")

pos2 <- apply(pmd1$u, c(1,2), function(x) return (max(x,0)))
neg2 <- apply(pmd1$u, c(1,2), function(x) return (min(x,0)))

par(mar=c(12,2,2,1))
barplot(t(pos2), col=2:3)
barplot(t(neg2), col=2:3, add=TRUE)
```

## Recreate factor patterns from wavelet transform

```{r echo=TRUE, eval=TRUE}
W <- GenW(n=32, filter.number=3, family="DaubExPhase")
idwt <- W[,-1] %*% pmd1$v;
plot(idwt[,1], type="l")
plot(idwt[,2], type="l")
```

## Applying Normal Topic Model 

```{r echo=TRUE, eval=TRUE}
K <- 2;
system.time(out2 <- NTMfit(wave_data, K=2));

barplot(t(out2$omega_out),col=2:(K+1),axisnames=F,space=0,border=NA,main="",las=1,ylim=c(0,1),cex.axis=1.5,cex.main=1.4)
title(main=paste("Structure Plot topic proportions,k=",K))
```

## Recreate cluster patterns from wavelet transform

```{r echo=TRUE, eval=TRUE}
W <- GenW(n=32, filter.number=3, family="DaubExPhase")
idwt <- W[,-1] %*% t(out2$alpha_out);
plot(idwt[,1], type="l")
plot(idwt[,2], type="l")

```

## Conclusions

The factors obtained did show bumpy patterns, but were not sinusoidal looking. Patterns consistent with and without wavelet transform.

